---
title: "A (more) technical `elisr` companion" 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A (more) technical `elisr` companion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`elisr` is a shorthand for exploratory likert scaling (ELIS) in R. ELIS builds
upon the multiple scaling approach and operates on a bottom up selection process
that integrates characteristic values of the classiscal test theory. This
technical companion is a step by step approach on how to use ELIS in R. At the
end of this manual, you'll have gained enough understanding of `elisr` to
confidently perform analysis in this framework. By working through the following
examples, you will not only grasp what's going on under the hood, but also
encounter some benefits of this procedure. Note however, that this is mainly
thought as a manual. For a more complete view on the heuristic potential that
lays out with this approach, you might want to take another look in the cited
paper as well. There is at least one article in prgress which is not mentioned
right now. I will update the list as soon as possible. All right, let's
summarize the introduction with the most important features. After that we dive
deeper into the technical details.

#### Most important features 

* The packages offers a multiple scaling approach according to the principle 
  of crystallization for (quasi-)metrical data in R.

* It thus supports you in exploring multidimensional data structures. 

  Note: In common research practice one generally considers explorative factor
  analysis in this case often accepting overly rigid model assumptions (e.g.,
  orthogonality).

* `elisr`, or rather its underlying algorithm, tries to bypass such inflexible 
  assumptions in a bottom up manner. 

* As a result the package provides a platform to maximize (a) the identification 
  of relevant dimensions, (b) allowing the scales to naturally correlate with 
  one another to accurately meet the actual conditions. 

 Hint: After exploratorily analyzing the data set with `disjoint()` and 
 `overlap()` you need to further process the result. Additional analysis in 
 mandatory. A place to start over is mentioned in the last section.
 
## `disjoint()` &  `overlap()`

`elisr` basically consists of two user functions `disjoint()` and `overlap()`.
With a typical case in mind, the practical difference between them is: (1)
`disjoint()` is set up to produce sharp and disjoint scale fragments. Sharp and
disjoint are those fragments, that include items which (a) share a strong linear
relationship with one another but where (b) any of them is tied to a single
fragment. That's where (2) `overlap()` comes into play. Passing fragments to
`overlap()` the functions underlying algorithm tries to enrich each fragment.
The emerging scales are fortified with items from your specified data frame but
those in the given fragment. Later on, we'll talk about the inclusion criterion
in greater detail. Making a long story short: Using `overlap()` an item can
appear in more than one of the enriched fragments. In doing so, we overcome the
splitting effect of the data frame induced by `disjoint()`. These basical
principles will unfold in step at a time throughout the rest of this companion.

Note however, that the paragraph above already reveals the key strategy of any
exploratory analysis using `elir`'s bottom up item selection procedure: (1)
Instruct `disjoint()` to produce a couple of sharp, disjoint scale fragments
from your specified data frame. (2) Let `overlap()` pick up the pieces; i.e.
enrich the disjoint fragments considering those items that are highly
correlating with the (sum score of the given) fragment. A more detailed
explanation goes along with the following examples.

### Analyzing the trust items using disjoint() and overlap()

Let's jump right into practice. `trust` is going to be the data base in the
upcoming units. `trust` is a subset of the German General Social Survey (ALLBUS)
2018 in which participants answered some questions on their trust in public
institutions and organizations. Type `?trust` if you want to get more
information. Because the data frame is already part of `elisr`, accessing it,
is pretty easily.

```{r}
library( elisr ) ; data( trust ) ; head( trust )
```

All right, let's get the machinery working. In the first snippet, we'll instruct
`disjoint()` to produce a couple of sharp fragments. Do that by setting a
relatively high value for the marginal corrected item total correlation
`mrit_min`. To cut things short, I'll use `.55` (mostly because this value
produced satisfactory results in prior analyses -- a more detailed
explanation on how to use this value you will find below).

```{r echo=TRUE}
# `(foo <- baz)`: assign and print in one step
(msdf <-  disjoint(df = trust, mrit_min = .55))
```

#### disjoint()

The table you could see in the output shows an object of type `msdf`. A `msdf`
is nothing more than a named `list` which I extended with a couple of attributes
for internal computational reasons (type`attributes(msdf)`). If you don't see
anything, something went wrong. I made `disjoint()` (and `overlap()`) actually
complain about lots of things -- and set up mostly sensible message for common
errors. My general advice is to first watch out for typos. These little goblins
are nasty everyday companions in applied research. But if you can't encrypt the
error message, hack it into google. Done? Alright, then we can go over to the
output section and a get a first indepth impression of `disjoint()`s machinery.

```{r}
msdf
```
Before we get started, we should learn some terminology. Therefore, focus on
`$scl_1`. I will refer to an objects like `scl_1` generally as a fragment. Now
look at (`eupalmnt, eucomisn`). This comma-separated pair of two items in the
first line, is the core of `scl_1`. The core marks the start of the upcoming
sclaing procedure -- a two item scale. As we could see, the core of `$scl_1$
includes the (lineraly) highest positivly correlating pair chosen from the data
frame you inserted (`trust`). To find it, `disjoint()` (as well as `overlap()`)
set up a correlation matrix and asks for the one that is the highest of all.

Technial note: If there are two equal correlations`disjoint()` will always
chooses the first one in the series. And that's how the inital of the scaling
procedure ends, by finding the highest correlating pair.

Tech tip: Neither `disjoint()` nor `overlap()` search the highest correlating
pair at any costs. The linear relationship between the core items must be
greater than your pre-set `mrit_min`. Thats the inclusion criterion I mentioned
above. However, I implemented a check for that, so `disjoint()` will let you
know, if this is not the case. Here, everything went well. Crosscheck that
result by looking at the first line: $mrit=rbar=0.92$. Remember; the
pre_specified value for the analysis was $mrit=.55$.

```{r}
msdf
```

After `disjoint()` started the two point item scale it tries to affiliate
another trust variable. In this case it picked up `polpati`. The algorithmic
background on this sequence is straightforward. First, `disjoint()` starts to
add up both items elementwise. This creates a sumscore. Subsequently, it watches
out for the item with the highest correlation with that sum (`polpati`). Because
the correlation of`polpati` with the added core items is still greater than our
specified stop criterion ($mrit_min=.55$) `polpati` is melted into the core. The
new fragment is now a bundle of 3 Items: `eupalament`, `eucomisn`, and
`polpati.` The same is true for `fedgovt` and reveals the important second step
of the algorithmic procedure: Continue to collect new items from `df` (here:
`trust`), until the correlation between the sum score of the items in the
fragment and any other item is less than your pre-specified `mrit_min.` In that
manner it soaks up the remaing three items you could see in the output -- namely
`bundtag`, `judsyst`, `fccourt`.


```{r}
msdf
```

Let's now focus on the question, why there is another scale with a `mrit_min`
value greater than the pre-specified one. The reason for this is that the
algorithm tries to build (and enrich) new fragments as long as there are more
remaining variables in your variable list (`df`). More specifially, the
algorithm won't stop to enrich a fragment as long as there is exactly one
variable left (in `trust`). Additionally, it won't interrupt constructing
fragments as long as there are at least two leftovers. But be aware of the fact
that again, `disjoint()` won't necessarily generate (and enrich) new fragments
at any costs. The correlations between items must be positive and greater than
your pre-specified minimum value (inclusion criterion). That betrays the last
unique step of this disjoint scaling procedure: If there are variables left that
meet the requirements, start another two point item scale. The result is once
more shown below.


```{r}
msdf
```

To sum up, you could test your understanding against the following question: Why
is there no other variable in `$scl_2`? Right, although there are 13 variables
in the data frame, there must have been no other variable among them that meets
the internally defined inclusion criterion -- namely, that the sumscore of the
whole fragment is greater than the pre-set $mrit_min$. As a consequence, no
other variable was considered to be merged with `{eupalmnt, eucomisn}`. Double
check in the last line of the given output.

```{r}
msdf
```

Let's tie in with above in a last step and zoom out on both fragments now. If we
remind ourselves on the description of `disjoint()`s task (to build disjoint
scale fragments), one can derieve from the last couple of steps, that it really
did its job. In the output each variable (`euparlament,...,tv`) is present in
either `$scl_1` or `$scl_2` exactly once. To put it another way, there is no
intersecting variable which overlaps in these two fragments. To assure that, I
made `disjoint()` exclude each variable after assign it to a given scale
fragment. In advance of the upcoming sections keep this result in mind.

#### overlap()

Before we'll actually try to enrich the disjointly built fragments, we should
reprint the primary output to easily catch up. A one-liner is sufficient for
this purpose. This one however is a little different. As you could see I
expicitally wraped the expression in `print()`. In the little R side note I will
explain why.

```{r}
print(msdf, digits=3)
# Pint output to n decimal places (default=2)
# print(msdf, digits=2)
```

A little R side note: When calling `msdf` R internally wraps the expression in
`print()`. Thus, `msdf` is a shortcut for `print(msdf)`. But since I wrote a
custom function to print objects of the eponymous type `print()` actually passes
the work on to `print.msdf()`. To get to the point, `print()` owns a nice way to
print the result to a specified number of digits, I equiped `print.msdf()` with
that either. The default is set to 2. Be aware of that omitting tons of decimal
places does not mean that the result is acurately rounded to the specified ones.
Anyway, I decided to avoid any rounding in the results. The reason for that is
R's (as I find it, deeply confusing) rounding standard  -- IEC 60559 (see
`?round`). So, if you get into the situation where you need to report more than
two decimal places, expand the result using `digits=n` and choose whatever
standard you like.

Let's resume and try to extend the scale with all items, despite those that are
already part of the result. To tie in to above, we're planning to resolve
`disjoint()`'s disjointness now. `overlap()` has the following role; it
(re-)considers the use of each item for a given fragment `disjoint()` has not
(simply because of its disjoint nature). The following two paragraphs will lay
out this ideas in much greater detail. Let's put the results on the desktop
first.

```{r}
(mosdf <- overlap(msdf, mrit_min = .4))
# Equivalent formalization unmasking the defaults
# (mosdf <- overlap(msdf, mrit_min = .4), overlap_with="fragment" )
```

A quick overview reveals that `overlap()` added a couple of variables to the
basic fragments from `disjoint()`. But what happend inside the machinerie? Well,
first `overlap()` finds all items in the data frame that are not part of the
given fragment. Let's call this its counterpart. Think about the counterpart as
items in the specified data frame minus the items within the fragment.
Accordingly for `$scl_1` the counterpart is made up of the following eight
items: `newsppr`, `tv`, `healserv`, `munadmin`, `tv`, `newsppr`, `uni` and
`police`. The second step is merely a repetition. `overlap()` starts the
previously described scaling procedure in each case over again. It thus
continuesly enriches each fragment with the according items from the counterpart
as long as the correlation between the sum score of all items and any other item
is less than the specified `mrit_min` (reminder:`mrit.min=.4`). To put it
another way, `overlap()` take the disjoint fragments as its basis when trying to
extend any of them. That's why the default option is called
`overlap_with="fragments"`. Additionally, `overlap()` provides the option to
choose only the highest positively correalting pair of each scale fragment
(`overlap_with="core"`). Type `?overlap` for more details. 

```{r}
mosdf
```

To let the key insights sink in, we should rework the whole process in a more
practical example once more. Have a look on the result of the previous function
call (`mosdf`). What you could see is that there are two fragments `$scl_1`and
`$scl_2`. At second glance they are pretty similar to the result of `msdf`.
Let's take $scl_1$ as an example.

```{r}
lapply( list( msdf=msdf$scl_1, mosdf=mosdf$scl_1 ), colnames )
# lapply( list( msdf=msdf$scl_2, mosdf=mosdf$scl_2 ), colnames )
```

That snippet shows the basis that `msdf` sets up through `disjoint()` for
`mosdf`. Likewise, one could say it demonstrates how `overlap()` extended this
fundament. Anyway, `overlap()` soaked up 6 additionl items from its counterpart
(`police`, ` newsppr`, `tv`, `uni`, `munadmin` & `healserv`). What happend here
algorithmically is, that `overlap()` searched for those items in `trust` that
were missing in `$scl_1`. Then, It used their intersection to build up this
counterpart. Subsequently, `overlap()` checked which correlation of these
variables with the sum score of the basic fragment is the greatest. If that
correlation exceeds the pre-set `mrit_min` it merged that variable into the
fragment. Finally, it starts over again considering the next item if it meets
these conditions. To conclude in more technical terms, those six additional
items tag the additional benefit we get from dissolving "`disjoint()`'s
genuin disjointness". 

```{r}
mosdf
```

There is one curcial step missing with regard to `$scl_2`. When `overlap()`
can't find any further item within a fragment that is worth considering it
continues; in this case `$scl_2`. So finally, what `overlap()` really is (well,
that's how I implemented it), is an enhancement of the disjoint scaling approach
by applying its underlying algorithm stepwise to multiple fragments considering
the according counterpart for their extension. This is what we see looping back
to the result of `overlap()`. There are two scaling fragments where some of the
items are present in both of them. I'll come to that in a moment. Let's clarify
the reason for that first. Even if a particular item is already part of $scl_1$
but nevertheless meets the pre-set admission criterion in `$scl_2`(the sum score
of the fragment and the corresponding item is greater $mrit_{min}=.4$) why not
tagging it as relevant for the developing scale? Note that `overlap()` will.

```{r}
mosdf
```

Now we can unfold the key mechanism when setting `mrit_min`: The more fragments
`disjoint()` produces the more reason one'll give `overlap()` to soak up
appropriate items from the corresponding counterparts. The key insight from
resolving `disjoint()`s disjointness (i.e. using `overlap()`) is that the scales
replicate. Check it against the result or type the follwing lines.

```{r}
lapply( list( scl_1=mosdf$scl_1, scl_2=mosdf$scl_2), 
        function(scl) sort(names(scl)))
```

To sum up, even if we crush the list of items when using `disjoint()` we allow
`overlap()` to pick up -- and expand -- its pieces. That is one of the
advantages of this approach. It don't force these fragments to be different.
Although it smashes the induced list of variables, it allows to explore how the
individual parts reuinte. Forcing these scale fragments to be different would
have masked the replication discovery itself. Hence, the breaking and reuiniting
part of this algorithmic procedure (namely the combined use of `disjoint()` and
`overlap()`) are inseparably connected. For this reason, we should consider
their use in a single call.

#### An all in one snippet 

```{r}
msdf <- overlap(
  disjoint( df = trust, mrit_min = .55),
  mrit_min = .4
) 
print(msdf, digits=2)
```

### Further anlysis of the trust items -- including negative correlations

In everyday research, one will usually stumble upon reversed items. These items
are broadly used to diminish response bias and reveal themselves in gerneral
through a negative correlation in the correlation matrix. Since there are no
reversed variables in `trust` we need to make up an artifical one.

```{r}
ntrust <- within( trust, tv <- 8 - tv )
# Compare the results 
# cor(trust, use="pairwise.complete.obs")
# cor(ntrust, use="pairwise.complete.obs")
```

What this code does, is to subtract $8$ from each value stored in the `uni`
variable, reassign it and store the result in a new data frame called `ntrust`.
From here we could start all over again using the all in one snippet. The only
thing we have to add is to set `negative_too=TRUE`and specify the start and
endpoint of the scale as a two element vector `c(1,7)`.

```{r}
ntrust <- within( trust, uni <- 8 - uni )
```

The first thing we should aim for is to replicate the results from the previous
analysis to understand the machinery and then soften the regularations to become
aware of the gained flexibility. But let's do this in a smoothed pace one step
at the time. To repeat the result from `disjoint()` we need to kind of
re-reverse `uni`. `disjoint()` does this for you, if you set the appropriate
options (`negative_too` & `sclvals`). Because we know from the prior results
that `uni` is not part of the fragments, it won't be re-reversed using
`disjoint()`. Usually we don't have this luxury. So setting `negative_too=TRUE`
here, seems like the right thing to do. Oh yes, for that reasons I made both
functions act very coperatively. `disjoint()` and `overlap()` will both let you
know, if they reversed an item (and if so, which one).

```{r}
(d <- disjoint( ntrust, mrit_min=.55, negative_too = TRUE, sclvals = c(1, 7) ))
```
The output is as expected. The result is identical to the one of the prior
analysis (using `trust` and `disjoint()`). We could now simply stick the result
we just stored in `d` into `overlap()`. Note however, that you have to tell
`overlap()`as well, that it should also consider to negative correlations. To
save you some typing, I made `overlap()` remember the start and endpoint of the
scale you've passed to `disjoint()` (see `attributes(d)$sclvals`). So no need to
specify them twice. What we should expect `overlap()` to do next, is to apply
the algortihm to trust therby reverse `uni` internally and second let you know. 

```{r}
overlap(d,
        # Note: overlap() remembers the scaling values from disjoint()
        mrit_min = .4, negative_too=TRUE
)

```
### Another made up (but autodidactic) example

Now its time to play around with the algorithm yourself. In the following
snippet I reversed a core item of the second fragment. Make predictions of what
will happen and why. Write them down and than and check the result against the
actual output. Done? Well, than remove `#` and see if both match.

```{r}
ntrust <- within( trust, tv <- 8 - tv )

# overlap( 
#  disjoint( ntrust, mrit_min=.55, negative_too = TRUE, sclvals=c(1,7) ),
#        mrit_min = .4, negative_too = TRUE)
```

### Different types of scales

There are 3 different types of scales which `disjoint()` and `overlap()` can
handle. They all have in common that the underlying variables are real numeric
vectors in R (either integers or objects of type `double`, see `typeof(x)`). If
you enable the `negative_too=TRUE` option `elisr`s wizards process:

1. Numeric scales starting at 1 (e.g., "1 2 3 4 5 6 7"). Both function use the
formula $(\max sclval + 1) - x$ when setting `scalvals=c(1,7)` in this case.
 
2. Numeric scales starting at 0 (e.g., "0 1 2 3 4 5 6 7). The use the formula
$\max sclval - x$. Set `sclvals=c(0,7)`. 

3. Numeric scales starting below 0 (e.g., "-3 -2 -1 0 -1 -2 -3"). The workhorse
here bases simply on the formula `x * (-1)`.

Those 3 option should cover a wide range of applications. Keep in mind, however,
that the various input possibilities do not stop with the abovementioned
examples. The only thing that might hold you back is defined by the logic of the
reversing rule itself. But within that range, you are free to enter whatever you
like.

### Handle items with different start- end endpoints

To be clear, `elisr` offers no way to tackle the start-endpoint issue
internally. You have to face and overcome this obstacle yourself. But it is
worth mentioning, that neither `overlap()` nor `disjoint()` will complain. They
silently obey and apply the algorithm to the given list of variables.

### Set up your own fragment

Sometimes, you'll have a concreat idea of how your fragment should look like. In
this case, you want to pre-determine the fragment and let `overlap()` soak up
additional variables afterwards. Prespecifying the variables that form your
fragment is straight forward. However, they differ slightly to the steps
performed so far. If you want to predetermine a fragment, just pass the reduced
version of your data frame to `disjoint()` -- but now use `mrit_min=0`. The
second crux move: overwrite `disjoint()`s data frame attribute. Set it to the
full list of variables you want do the overlap with. Everything else stays the
same. Hand the modified object over to `overlap()` and that's it.

```{r}
frag <- trust[c("tv", "bundtag", "fccourt")]
pre <- disjoint(df = frag, mrit_min = 0)
# overlap() uses this attribute to build the counterpart
attributes(pre)$df <- trust
(msdf <- overlap(pre, mrit_min = .4))
```
Technical side note: If you want to know why this additonal step is necessary,
we need to get a little more into technical details. By running `disjoint()` on
a subset of trust items, the function memorizes `frag` from your call as its
data frame attribute (see `attributes(pre)$df`). The moment you commission
`overlap()`, the function accesses `disjoint()`'s attributes. In this case
`overlap()` tries to evaluate `df=core`. Why? Remember that `overlap()` needs an
idea of the variables it should considers for the upcoming extension. To find
inspiration, it evaluates the `df` argument and seperates the items within the
given fragment from those left in the data frame. But because the variables in
the fragment are equal to those in the data frame, its counterpart is empty. In
other words, there are no items to extend the fragment with. If we set this
attribute manually however, we sneak in a couple of new item and therefrore
induce `overlap()` to grab items from this object -- despite `tv`, `bundtag` and
`fccourt`. 

Additional warning: Please be aware that this is an advanced application. It is
a serious change in `elisr`s mechanism, because you bypass a great bunch of
internal security measures. So assure that at least the predetermined variables
are a real subset of the data frame you've planed to assign as an attribute.

## Working with the results

In the previous chapters I mentioned the `print.msdf()` function. Remeber, the
idea was to exploit the fact, that R internally wraps `msdf` in `print()` and
intervene in the process. `print()` now passes the work on to the `print.msdf()`
which presents the result in the following manner:

```{r}
msdf <- overlap( 
        disjoint( ntrust, mrit_min=.55, negative_too = TRUE, sclvals=c(1,7) ),
        # Note: overlap() remembers the scaling values from disjoint()
        mrit_min = .4, negative_too = TRUE 
)
```

The question is, why am I telling you all this? Well, because what you see is
not the actual result of `disjoint()` or `overlp()`, but a processed result by
`print.muscldf()`. To put it another way, the actual outcome is hidden by this
internal printing mechanism. But you can break through this process by adressing
one of `msdf`s elements directly. This reveals the internally sorted (and in
this case partially reversed, plus extended) data list. To keep things
organized, the following output shows only the first six values (`head(x)`) of
`msdf`'s component one: `$scl_1`.

```{r}
head(msdf$scl_1)
```

There is one tiny thing I want to point out here, before we dive deeper into the
topic of this section. Look at `tv`. It is reversed now. Thus, the reversing
process was successful. If you want to prove that, just compare the outcome to
prevoius results in this section and you'll see that we could actually
replicated them (type: `overlap(disjoint(trust, mrit_min=.55), mrit_min=.4`). In
a nutshell: `disjoint()` and `overlap()` both keep the reversed form of the
variable. This behavior is thought to make further analysis as convenient as
possible. Let's move on by discussing why.

Because further anaylsis of the result is mandatory, I made the functions
translate the output into a object type most connecting packages can handle.
That is a `data.frame`. So, inside the machinery, specifying `scl_1` in `msdf`
breaks through the inherent structure and transfers the result to a data frame
object.

```{r}
class(msdf$scl_1)
```

To make a final unequivocal statement on `elisr`s results. What you really get
from applying `disjoint()` and `overlap()` is an intermediate. A good hint is
`mrit`. `mrit` is again the /marginal/ corrected item total correlation.
Remember, what that means: The linear relationship between the sum score of the
items in a fragment at some specific point of the scale development process and
the item which is considered for admission. Hence, the output represents the
construction from a bottom-up perspective. To put it another way, we gradually
build scales from scratch and `mrit` as well as `rbar`, and `alpha` monitor this
procsess for you. But this result however can differ significantly from a more
comprehensive (reliability) analysis of a proposed scale. Therefore I'll end the
manual with a place for you to (re-)start.

## A place to restart 

There are several options to continue. But the one which I find most appealing,
is part of the `psych` package. The object of desire is called `alpha()`. This
neat function provides a bunch of helpful diagnostic information on the proposed
scale(s). The snippet below checks if psych is present and give advise on how to
procced. Just copy and run it.

```{r}
if (requireNamespace("psych", quietly = TRUE)) {
  cat("`psych` is present. Ready to go!\n")
} else {
  cat("Please install the psych package to continue, type:\n")
  message("install.packages('psych')")
}
```
Once you have installed and loaded `psych`, there are two things you need to do.
First, store the result of an analysis into an object. Second, hand the scale
over to `alpha()` That's basically it. One scale is already available through
`msdf`. Let's take that. Note that `alpha()`s output is pretty extensive.
Therefore, I suspressed the result. But if you run the code yourself, you will
see the result in full length.


```{r echo=T, results='hide'}
psych::alpha(msdf$scl_1)
```

As we've seen throught the manual, an analysis with `elisr` often provides a
couple of suggestions. We should work out a solution for this scenario, too. I
usually hack the result of `overlap()` in `lapply()` and let `alpha()` do it's
job on all components of that modified list. Let's do that. But before we
start, please note once more, that the full result will be present only if you
run the code yourself.
 
```{r echo=T, results='hide'}
msdf <- overlap(d, mrit_min = .4)
lapply(msdf, psych::alpha)
```
