---
title: "Introduction to `musclr`" 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{musclr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(musclr)
```

`musclr` is a shortcut for "multiple scaling according to the principle of 
crystallization in R". The vignette is a step by step approach to teach, first 
and foremost, how to use `musclr`. In doing so (i.e., following the examples 
below) you will implicitly encounter some benefits of this exploratory 
procedure which primary lead to the development of this package. But if one
is ought for the full range of advantages, he or she might have a second look in 
the referenced paper. Let's summarize the introduction with the most important 
features.

#### Most important features 

* `musclr` offers a multiple scaling approach according to the principle 
  of crystallization for (quasi-)metrical data in R.

* It thus supports you in exploring multidimensional data structures. 

  Note: In common research practice one generally considers explorative factor
  analysis in this case often accepting rigid model assumptions (e.g.,
  orthogonality). 

* `musclr`, or rather its underlying algorithm, tries to bypass such inflexible 
  assumptions. 

* As a result the package provides a platform to maximize (a) the identification 
  of relevant dimensions, (b) allowing the scales to naturally correlate with 
  one another to accurately meet the actual conditions. 

 Hint: After exploratorily analyzing the data set with `disjoint()` and 
 `overlap()` you need to further process the result. Additional analysis in 
 mandatory. A solution on how to proceed is mentioned in the upcoming sections. 
 
# `disjoint()` &  `overlap()`

`musclr` basically consists of two functions: `disjoint()` and  `overlap()`. 
 With a typical case in mind, the practical difference between them is:
 (1) You will mainly use `disjoint()` to produce "sharp" but disjoint scale 
 fragments from your data frame. What that means, is items in this fragment 
 share a strong linear relationship but any of them belongs only to a single 
 fragment. (2) Passing these fragments to  `overlap()` accordingly, will "enrich" each of them, by trying to attach an 
 item  
 
That reveal the key vorgehensweise of any analysis: Anleiten disjoint to produce
a high number of "sharp" disjoint scale fragments from your data frame and let
overlap collect an


a high value in order to get
sharp scale fragments and th
 



##

Mokken 
 
 
Mokken, R. J., 1971: A Theory and Procedure of Scale
Analysis. With Applications in Political Research. The
Hague/New York: Mouton.
 
 
 
 
Note: devalere the mjmisc (?) package in /Description since otherwise the package
doesnt run on CRAN 


