---
title: "`musclr` -- Multiple Scaling According to the Principle of Crystallization in R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{musclr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(musclr)
```

# An opener 

The vignette is a step by step approach to teach, first and foremost, how to 
use the `musclr` package. In doing so one will automatically encounter a couple 
of benefits benefits of this procedure which lead to the development of this 
package. Let's start with the most important features.

## Aims, goals and expectations

* `musclr` implement the multiple scaling approach according to the principle 
    of crystallization for _metrical_ data in R.
    
* It supports the user when exploring multidimensional data structures. 
 
   Note: In common research practice one generally considers 
   explorative factor analysis in this case -- often accepting rigid model 
   assumptions (e.g., orthogonality). 
    
* The packages, or rather its underlying algorithm, tries to bypass 
  inflexible assumptions. 
    
 * As a result `musclr` provides a platform to maximize (a) the identification 
   of relevant dimensions, (b) allowing the scales to naturally correlate with 
   one another to meet the actual conditions. I made a suggestion ... advice
 
## Concluding note 
  
 After exploratorily analyzing the data set with `disjoint()` 
 and `overlap()` one need to further process the result. Additional packages, 
 i.e., further analysis is mandatory. 
  
 
 
 
 
Note: devalere the mjmisc (?) package in /Description since otherwise the package
doesnt run on CRAN 


